Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 2.dev

### 2024-10-09 main.tex
Shortened header and mentioned Frans role in funding.
Final pass of edits just before submitting.

\abstract{
  Uncontrolled bicycles are generally unstable at low speeds. We add an
  automatically controlled steering motor to a consumer electric bicycle that
  stabilizes the riderless bicycle down to about 3~\si{\kph}. We hypothesize
  that a stabilized bicycle will reduce the probability of falling. To test the
  motor's possible assistance during falls, we apply varying magnitude external
  handlebar perturbations to twenty-six participants who rode on a treadmill
  with the balance assist system both activated and deactivated. The
  probability of recovering from a handlebar perturbation significantly
  increases when the balance assist is activated at a travel speed of
  6~\si{\kph}. This positive effect is most prominent at and around the
  individual riders' perturbation resistance threshold. We conclude that use of
  a balance assist system in real world bicycling would reduce the number of
  falls that occur near riders' control authority limits.
}

@@

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage of reported serious injuries~\citep{Wegman2024}. At low speeds, from
start up to typical cruising speeds, bicycles are not self-stable and can be
challenging for the rider to balance. Low speed crashes may be reduced if the
bicycle was self-stable at these speeds by relieving the rider of some of their
necessary control activity. Bicycles can be mechanically modified to lower the
speeds at which they are self-stable~\citep{Astrom2005} and since the 1980s it
has been known that applying a motor actuated steering torque proportional to
the vehicle's roll angular rate~\footnote{Roll angular rate is defined herein
as per the SAE J670 vehicle standards.} can stabilize a single-track vehicle
down to very low speeds~\citep{Ruijs1986}. If automatic control of steering can
stabilize a bicycle, it may reduce the control required from the rider to
successfully manage balancing tasks much like the natural high speed
self-stability already does. We have developed a balance-assisting bicycle,
Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack first introduced in \citet{Alizadehsaravi2023}.}
  \label{fig:balance-assist-bicycle}
\end{figure}

Riders fall when their control authority is not able to maintain a stable
vehicle state. There are many real world scenarios that can put the vehicle
into an uncontrollable state. External forces applied to the vehicle or rider
are one such scenario type and natural examples include wind gusts, handlebars
colliding with a neighbor's, a bag swinging from the handlebar, or simply
hitting a bump in the road. To assess our balance assisting bicycle, we subject
the rider to perturbations at the handlebar, which can relatively easily cause
a rider to fall.

@@

\subsection{Technical Background}
%
During the early years of developments in automatic control,
\citet{Whipple1899} not only derived the correct equations of motion of the
bicycle \citep{Meijaard2007} but realized and showed that roll motion feedback
can stabilize a bicycle. Much later, attempts at automatic roll stabilization
of single-track vehicles began in earnest after predictive motorcycle models
were developed and refined throughout the 1970s. \Citet{vanZytveld1975} was
influenced by Whipple's work and seems to be the first to attempt to
robotically stabilize a small motorbike with a controlled inverted pendulum
that mimicked rider lean, but he was not successful in demonstrating what his
theoretic control model correctly predicted. In his model, he recognized that
feedback of the vehicle roll angle and angular rate was essential to stabilize
the vehicle. It was not until the early 1980s that \citet{Nagai1983}
successfully demonstrated balancing a robotic bicycle on a treadmill with both
steering control and a laterally moving mass. \citet{Ruijs1986} followed this
breakthrough by demonstrating an automatically balanced motorcycle and they did
so solely with a steering motor. Ruijs and Pacejka showed that steer torque
driven by roll angle feedback stabilizes the capsize mode, by roll angular rate
feedback stabilizes the weave mode, and by steer angular rate feedback
stabilizes the wobble mode.~\footnote{These motorcycle (and bicycle) eigenmodes
are defined in \citep{Sharp1971}.} They also showed how the control gains must
change with respect to vehicle speed for favorable control across all speeds.
This roll motion feedback enables the simplest controller that can stabilize a
single-track vehicle above a minimum speed when one is not concerned with
wobble instabilities. But Ruijs and Pacejka's work was not particularly
concerned with low speed stability and their vehicle was fully automatic, i.e
no human rider was involved.

Many more automatically balanced single-track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in human balancing and what effect it may have towards
single-actor falls. Most of these robotic bicycle and motorcycle designers did
not intend for a human rider to also control the stabilized vehicle.
Nevertheless, an automatically stabilized bicycle can be controlled by a human
rider if the motor controlled steer torque and the rider applied steer torque
act on the steer in parallel. The effect of this automatic control gives the
ability to effectively change the human-controlled plant dynamics, up to some
limits. In our prior study~\citep{Alizadehsaravi2023}, we demonstrate reduced
motion during distractions due to the balance assist system but
\citet{Hanakam2023} recently showed rider dissatisfaction with the
stabilization of a similar vehicle so the possible benefits are not
definitively established.

The design of a balance assist system relies on the linear Carvallo-Whipple
bicycle model~\citep{Carvallo1899,Whipple1899} which is the simplest bicycle
model that exhibits non-minimum phase behavior and self-stability. The bicycle
model is well suited for showing the effect of a roll motion feedback driven
steer motor on the dynamics. This model is equivalently valid for on-road or
treadmill riding~\citep{Kooijman2009a}, which have the same fundamental
dynamics. The linear version of this model can be described by the fourth order
state space equations:
%
\begin{align}
  \dot{\vec{x}} = \mathbf{A} \vec{x} +
  \mathbf{B}
  \vec{u}
  \textrm{ where }
  \vec{x} = \begin{bmatrix} \phi \\ \dot{\phi} \\ \delta \\ \dot{\delta} \end{bmatrix}
  \textrm{ and }
  \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
  \textrm{.}
\end{align}
%
The states \(\vec{x}\) are the roll angle \(\phi\) and steer angle \(\delta\)
along with their time derivatives and the inputs \(\vec{u}\) are roll torque
\(T_\phi\) and steer torque \(T_\delta\). The state matrix \(\mathbf{A}\) is a
function of the equilibrium forward speed \(v\). It and the input matrix
\(\mathbf{B}\) are otherwise populated with expressions that are functions of
the geometric and inertial parameters of the nonholonomic multibody system made
up of four rigid bodies: two wheels, front frame, and rear frame.

@@

\section{Methods}
%
\subsection{Bicycle}
%
We modified a Grenoble C8 HMB electric bicycle (Royal Dutch Gazelle, Dieren,
The Netherlands) with a custom motor in the steering assembly capable of
applying up to 7~\si{\newton\meter} of torque between the head tube and steer
tube, see Figure~\ref{fig:balance-assist-bicycle}. A custom motor controller
converts the commanded torque to applied torque. We measure the speed of the
rear wheel with a DF30 Speed Sensor (Bosch eBike Systems, Reutlingen, Germany)
and measure the body fixed roll rate of the bicycle with a VR IMU BN0086 MEMs
rate gyroscope (Sparkfun, Niwot, USA). The balance assist control algorithm is
implemented on a Teensy microprocessor (PJRC, USA) and data from all sensors is
logged with a CanEdge2 CAN bus (CSS Electronics, Aabyhoej, Denmark) at least
100 Hz.

\subsection{Balance Assist Control}
%
We use a forward speed \(v\) gain scheduled proportional roll angular rate
feedback controller to stabilize the bicycle. In the speed range tested, the
commanded steer torque \(T^\textrm{m}_\delta\) from the steer motor follows the
control law
%
\begin{align}
  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = \kappa(v_{stable} - v)\dot{\phi}
  \label{eq:implemented-controller}
\end{align}
%
where \(v_{\textrm{stable}} = 4.7~\si{\meter\per\second}\) is approximately the
average stable speed predicted from the open loop bicycle rigid-rider dynamics.
We use the gain factor \(\kappa=8\) (low) and \(\kappa=10\) (high) to tune the gain
magnitude during the experiments. Scaling the proportional feedback gain
linearly with respect to speed stabilizes the normally unstable weave mode of
the bicycle down to 2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
4.6~\si{\kilo\meter\per\hour} for the ridden bicycle as shown in
Figure~\ref{fig:eigenvalues}.
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row) and controlled with \(\kappa=10\) (lower row)
    root locus of eigenvalue components (real: solid, imaginary: dashed)
    plotted versus speed for the bicycle without the inertial effect of a rigid
    rider (left) and with a rigid rider (right). Vertical dotted and
    dotted-dashed lines indicate the two speeds we perform experiments at. Grey
    shaded region is the linear stable speed range. Geometric and inertial
    parameters for these plots were estimated with the methods presented in
    \citet{Moore2012} and software packages
    BicycleParameters~1.1.1~\citep{Moore2024} and
    Yeadon~1.5.0~\citep{Dembia2015} and are shown in
    Appendix~\ref{sec:physical-parameters}.}
  \label{fig:eigenvalues}
\end{figure}

@@

The participants were divided into two groups. The first group of fifteen
participants performed the protocol at a belt speed of
10~\si{\kph}(2.8~\si{\mps}) with the controller factor set to \(\kappa=8\) and two
weeks later the second group of eleven particpants performed the protocol at a
belt speed of 6~\si{\kph} (1.7~\si{\mps}) with the controller factor set to
\(\kappa=10\).

@@

\subsection{Measurements}
%
We measure the time histories of the Bump'Em delivered perturbation forces and
the bicycle's steer angle \(\delta\), roll angle \(\phi\), roll angular rate
\(\dot{\phi}\), and forward speed \(v\). Figure~\ref{fig:perturbation_0} shows
an example perturbation force measurement compared to our Bump'Em controller
command.
%
\begin{figure}
  \centering
  \includegraphics[width=140mm]{figures/torque_angle_perturbation_10.png}
  \caption{Externally applied perturbation torque alongside the resulting
    measured motion and steer motor induced torque based on a 110~\si{\newton}
    counterclockwise applied force at a 6~\si{\kph} travel speed. The circles
    on the roll and steer angle plot show the angles \(\phi_0\) and
    \(\delta_0\) at the perturbation start.}
  \label{fig:perturbation_0}
\end{figure}

Based on findings from measuring riders without balance assist in
\citet{Haitjema2023a}, we calculate several variables that we hypothesize may
influence fall probability. We use the angular impulse \(L\) of the
perturbation forces over a 0.3~\si{\second} duration to characterize the
magnitude of delivered perturbation. The duration is selected based on the
duration of the commanded force and is calculated as follows:
%
\begin{align}
  L =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}(F_r + F_l) dt
  =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left[(F_{rf} - F_{rr}) + (F_{lf}
  - F_{lr})\right] dt
  \label{eq:angular-impulse}
\end{align}
%
where \(F_r\) and \(F_l\)  is the total force applied on the right and left
handlebar ends, respectively which are the sum of the rear and front load cell
readings \(F_{rr}\) and \(F_{rf}\), for example. The handlebar length is given
as \(l\) in Equation \ref{eq:angular-impulse}. We use angular impulse instead
of peak torque to normalize for the duration of the applied perturbation.

We record the order number of each perturbation \(j\), i.e. first, second,
third, \ldots, to measure how many perturbations a rider is subjected to before
the current perturbation. At the initiation of each perturbation we log the
instantaneous steer and roll angles to characterize the configuration of the
bicycle when perturbed. The gain factor setting on the balance assist controller
indicates if the assistance is off \(\kappa=0\) or on at two different levels: low
\(\kappa=8\) or high \(\kappa=10\). A recovery from the perturbation is successful if the
person neither places their foot down onto the treadmill surface nor their
wheel of the bicycle exits the width of the treadmill belt. We record this as a
binary variable \(f\) for ``fall''. All measured variables are reported in
Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall Outcome'' and
    ``Perturbation Order Number'' are recorded per perturbation instance. The
    remaining measures are time varying during each trial.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain Factor & \(\kappa\) & \si{\newton\second\squared} & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
    Fall Outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear &
    \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & inline load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & BN0086 rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

\subsection{Statistics}
%
We test our hypothesis that the balance assist controller will reduce the
probability of falling when perturbed externally at the handlebar. We have a
single binary fall outcome variable \(f\) that is dependent on several possible
explanatory independent variables, one of which is the binary balance assist
state (on or off). See Table~\ref{tab:stat-model-variables} for the statistical
model variables.
%
\begin{table}
  \centering
  \caption{Independent and dependent variables used in the statistical
    regression model.}
  \label{tab:stat-model-variables}
  \begin{tabular}{llll}
    \toprule
    Variable & Causality & Units & Description \\
    \midrule
    \(L\)        & independent & \si{\newton\meter\second} & angular impulse of perturbation torque \\
    \(\delta_0\) & independent & \si{\degree} & steer angle at start of perturbation \\
    \(\phi_0\)   & independent & \si{\degree} & roll angle at start of perturbation \\
    \(f\)        & dependent   & boolean & outcome: did not fall 0 or did fall 1 \\
    \(s\)        & independent & boolean & balance assist: off 0 or on 1 \\
    \(j\)        & independent & integer & order number of perturbation \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\section{Results}
%
The coefficient estimates for a single-level logistic regression for the data
from the 6~\si{\kph}, gain factor \(\kappa=10\) trials are shown in
Table~\ref{tab:freq-coefs-6}. The angular impulse, perturbation order, and
balance assist state are all statistically significant predictors with angular
impulse having a dominant effect. Larger angular impulse increases the
probability to fall and both enduring more perturbations or having the balance
assist on, decrease the probability to fall. The associated multiplicative
change in odds are shown in Table~\ref{tab:freq-coefs-6}.
%
\begin{table}
  \centering
  \caption{Logistic regression coefficient estimates \(\alpha_k\) at
    6~\si{\kph} and gain factor \(\kappa=10\) along with the standard error SE, p-value
    \(p\), multiplicative change in odds \(e^{\alpha_k}\), and the 5\%
    confidence interval bounds.}
  \label{tab:freq-coefs-6}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & \(\alpha_k\) & SE & \(p\) & \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
    \midrule
    Intercept \(\beta\)                           & -0.29 & 0.17 & 0.09 & 0.75 & 0.53 & 1.05 \\
    Angular impulse \(L\)                         & 1.69 & 0.27 & 0.00\(^*\) & 5.40 & 3.18 & 9.16 \\
    Perturbation order \(j\)                      & -0.77 & 0.22 & 0.00\(^*\) & 0.46 & 0.30 & 0.72 \\
    Balance assist state \(s\)                    & -0.64 & 0.27 & 0.02\(^*\) & 0.53 & 0.31 & 0.89 \\
    Roll angle \(\phi_0\)                         & -0.25 & 0.21 & 0.24 & 0.78 & 0.51 & 1.18 \\
    Steer angle \(\delta_0\)                      & -0.14 & 0.21 & 0.51 & 0.87 & 0.58 & 1.32 \\
    Balance assist state $\times$ roll angle         & 0.52 & 0.34 & 0.12 & 1.68 & 0.86 & 3.29 \\
    Balance assist state $\times$ steer angle        & -0.41 & 0.34 & 0.22 & 0.66 & 0.34 & 1.28 \\
    Balance assist state $\times$ angular impulse    & 0.41 & 0.41 & 0.32 & 1.51 & 0.67 & 3.38 \\
    Balance assist state $\times$ perturbation order & -0.53 & 0.34 & 0.12 & 0.59 & 0.30 & 1.15 \\
    \bottomrule
  \end{tabular}
\end{table}

>>>>

\abstract{
  Uncontrolled bicycles are generally unstable at low speeds. We add an
  automatically controlled steering motor to a consumer electric bicycle that
  stabilizes the riderless bicycle down to about 3~\si{\kph} to assist a rider
  in balancing the vehicle. We hypothesize that a such a stabilized bicycle
  will reduce the probability of falling. To test the system's possible
  assistance during falls, we applied varying magnitude external handlebar
  perturbations to twenty-six participants who rode on a treadmill with the
  balance assist system both activated and deactivated. We show that the
  probability of recovering from a handlebar perturbation significantly
  increases when the balance assist is activated at a travel speed of
  6~\si{\kph}. This positive effect is most prominent at and around the
  individual riders' perturbation resistance threshold. We conclude that use of
  a balance assist system in real world bicycling can reduce the number of
  falls that occur near riders' control authority limits.
}

@@

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage of reported serious injuries~\citep{Wegman2024}. At low speeds, from
start up to typical cruising speeds, bicycles are not self-stable and can be
challenging for the rider to balance. Low speed crashes may be reduced if the
bicycle was self-stable at these speeds by relieving the rider of some of their
necessary control activity. Bicycles can be mechanically modified to lower the
speeds at which they are self-stable~\citep{Astrom2005} and since the 1980s it
has been known that applying a motor actuated steering torque proportional to
the vehicle's roll angular rate~\footnote{Roll angular rate is defined herein
as per the SAE J670 vehicle standards.} can stabilize a single-track vehicle
down to very low speeds~\citep{Ruijs1986}. If automatic control of steering can
stabilize a bicycle, it may reduce the control required from the rider to
successfully manage balancing tasks much like natural high speed
self-stability already does. We have developed a balance-assisting bicycle,
Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack first introduced in \citet{Alizadehsaravi2023}.}
  \label{fig:balance-assist-bicycle}
\end{figure}

Riders fall when their control authority is not able to maintain a stable
vehicle state. There are many real world scenarios that can put the vehicle
into an uncontrollable state. External forces applied to the vehicle or rider
are one such scenario type and natural examples include wind gusts, handlebars
colliding with a neighbor's, a bag swinging from the handlebar, or simply
hitting a bump in the road. To assess our balance assisting bicycle, we subject
the rider to perturbations at the handlebar, which can easily cause a rider to
fall.

@@

\subsection{Technical Background}
%
During the early years of developments in automatic control,
\citet{Whipple1899} not only derived the correct equations of motion of the
bicycle \citep{Meijaard2007} but realized and showed that roll motion feedback
can stabilize a bicycle. Much later, attempts at automatic roll stabilization
of single-track vehicles began in earnest after predictive motorcycle models
were developed and refined throughout the 1970s. \Citet{vanZytveld1975} was
influenced by Whipple's work and seems to be the first to attempt to
robotically stabilize a small motorbike with a controlled inverted pendulum
that mimicked rider lean, but he was not successful in demonstrating what his
theoretic control model correctly predicted. In his model, he recognized that
feedback of the vehicle roll angle and angular rate was essential to stabilize
the vehicle. It was not until the early 1980s that \citet{Nagai1983}
successfully demonstrated balancing a robotic bicycle on a treadmill with both
steering control and a laterally moving mass. \citet{Ruijs1986} followed this
breakthrough by demonstrating an automatically balanced motorcycle and they did
so solely with a steering motor. Ruijs and Pacejka showed that steer torque
driven by roll angle feedback stabilizes the capsize mode, by roll angular rate
feedback stabilizes the weave mode, and by steer angular rate feedback
stabilizes the wobble mode.~\footnote{These motorcycle (and bicycle) eigenmodes
are defined in \citep{Sharp1971}.} They also showed how the control gains must
change with respect to vehicle speed for favorable control across all speeds.
Thus roll motion feedback enables the simplest controller that can stabilize a
single-track vehicle above a minimum speed when one is not concerned with
wobble instabilities. But Ruijs and Pacejka's work was not particularly
concerned with low speed stability and their vehicle was fully automatic, i.e
no human rider was involved.

Many more automatically balanced single-track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in human balancing and what effect it may have towards
single-actor falls. Most of these robotic bicycle and motorcycle designers did
not intend for a human rider to also control the stabilized vehicle.
Nevertheless, an automatically stabilized bicycle can be controlled by a human
rider if the motor controlled steer torque and the rider applied steer torque
act in parallel. The effect of this automatic control gives the ability to
effectively change the human-controlled plant dynamics, up to some limits. In
our prior study~\citep{Alizadehsaravi2023}, we demonstrate reduced motion
during distractions and light perturbations due to the balance assist system
but \citet{Hanakam2023} recently showed rider dissatisfaction with the
stabilization of a similar vehicle so the overall possible benefits are not
yet definitively established.

The design of our balance assist system relies on the linear Carvallo-Whipple
bicycle model~\citep{Carvallo1899,Whipple1899} which is the simplest bicycle
model that exhibits both the non-minimum phase behavior ``countersteering'' and
self-stability. The bicycle model is well suited for showing the effect of a
roll motion feedback driven steer motor on the dynamics. This model is
equivalently valid for on-road or treadmill riding~\citep{Kooijman2009a}, which
have the same fundamental dynamics. The linear version of this model can be
described by the fourth order state space equations:
%
\begin{align}
  \dot{\vec{x}} = \mathbf{A} \vec{x} +
  \mathbf{B}
  \vec{u}
  \textrm{ where }
  \vec{x} = \begin{bmatrix} \phi \\ \dot{\phi} \\ \delta \\ \dot{\delta} \end{bmatrix}
  \textrm{ and }
  \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
  \textrm{.}
\end{align}
%
The states \(\vec{x}\) are the roll angle \(\phi\) and steer angle \(\delta\)
along with their time derivatives and the inputs \(\vec{u}\) are roll torque
\(T_\phi\) and steer torque \(T_\delta\). The state matrix \(\mathbf{A}\) is a
function of the equilibrium forward speed \(v\). It and the input matrix
\(\mathbf{B}\) are otherwise populated with expressions that are functions of
the geometric and inertial parameters of the nonholonomic multibody system made
up of four rigid bodies: two wheels, front frame, and rear frame.

@@

\section{Methods}
%
\subsection{Bicycle}
%
We modified a Grenoble C8 HMB electric bicycle (Royal Dutch Gazelle, Dieren,
The Netherlands) with a custom motor in the steering assembly capable of
applying up to 7~\si{\newton\meter} of torque between the head tube and steer
tube, see Figure~\ref{fig:balance-assist-bicycle}. A custom motor controller
converts the commanded torque to applied torque. We measure the speed of the
rear wheel with a DF30 Speed Sensor (Bosch eBike Systems, Reutlingen, Germany)
and measure the body fixed roll rate of the bicycle with a VR IMU BN0086 MEMs
rate gyroscope (Sparkfun, Niwot, USA). The balance assist control algorithm is
implemented on a Teensy microprocessor (PJRC, Sherwood, USA) and data from all
sensors is logged with a CanEdge2 CAN bus (CSS Electronics, Aabyhøj, Denmark)
at least 100 Hz.

\subsection{Balance Assist Control}
%
We use a forward speed \(v\) gain scheduled proportional roll angular rate
feedback controller to stabilize the bicycle. In the speed range tested, the
commanded steer torque \(T^\textrm{m}_\delta\) from the steer motor follows the
control law
%
\begin{align}
  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = \kappa(v_{stable} - v)\dot{\phi}
  \label{eq:implemented-controller}
\end{align}
%
where \(v_{\textrm{stable}} = 4.7~\si{\meter\per\second}\) is approximately the
average stable speed predicted from the open loop bicycle rigid-rider dynamics.
We use the gain factor \(\kappa=8\) (low) and \(\kappa=10\) (high) to tune the gain
magnitude during the experiments. Scaling the proportional feedback gain
linearly with respect to speed stabilizes the normally unstable weave mode of
the bicycle down to 2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
4.6~\si{\kilo\meter\per\hour} for the ridden bicycle as shown in
Figure~\ref{fig:eigenvalues}.
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row) and controlled with \(\kappa=10\) (lower
    row) root locus of the eigenvalue components (real: solid, imaginary:
    dashed) plotted versus speed for the linear Carvallo-Whipple bicycle model
    without the inertial effect of a rigid rider (left) and with a rigid rider
    (right). Vertical dotted and dotted-dashed lines indicate the two speeds we
    perform experiments at: 6~\si{\kph} and 10~\si{\kph}. The grey shaded
    region is the linear stable speed range. Geometric and inertial parameters
    for these plots were estimated with the methods presented in
    \citet{Moore2012} and software packages
    BicycleParameters~1.1.1~\citep{Moore2024} and
    Yeadon~1.5.0~\citep{Dembia2015} and are shown in
    Appendix~\ref{sec:physical-parameters}.}
  \label{fig:eigenvalues}
\end{figure}

@@

The participants were divided into two groups. The first group of fifteen
participants performed the protocol at a belt speed of
10~\si{\kph}(2.8~\si{\mps}) with the gain factor set to \(\kappa=8\) and two
weeks later the second group of eleven participants performed the protocol at a
belt speed of 6~\si{\kph} (1.7~\si{\mps}) with the gain factor set to
\(\kappa=10\).

@@

\subsection{Measurements}
%
We measure the time histories of the Bump'Em delivered perturbation forces and
the bicycle's steer angle \(\delta\), roll angle \(\phi\), roll angular rate
\(\dot{\phi}\), and forward speed \(v\). Figure~\ref{fig:perturbation_0} shows
an example perturbation force measurement compared to our Bump'Em controller
command.
%
\begin{figure}
  \centering
  \includegraphics[width=140mm]{figures/torque_angle_perturbation_10.png}
  \caption{Example of the externally applied perturbation torque alongside the
    resulting measured motion and steer motor induced torque based on a
    110~\si{\newton} counterclockwise applied force at a 6~\si{\kph} travel
    speed. The circles on the roll and steer angle plot show the angles
    \(\phi_0\) and \(\delta_0\) at the perturbation start.}
  \label{fig:perturbation_0}
\end{figure}

Based on findings from measuring riders without balance assist in
\citep{Haitjema2023a}, we calculate several variables that we hypothesize may
influence fall probability. We use the angular impulse \(L\) of the
perturbation forces over a 0.3~\si{\second} duration to characterize the
magnitude of delivered perturbation. The duration is selected based on the
duration of the commanded perturbation force and is calculated as follows:
%
\begin{align}
  L =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left(F_r + F_l\right) dt
  =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left[\left(F_{rf} -
    F_{rr}\right) + \left(F_{lf} - F_{lr}\right)\right] dt
  \label{eq:angular-impulse}
\end{align}
%
where \(F_r\) and \(F_l\)  is the total force applied on the right and left
handlebar ends, respectively which are the sum of the rear and front load cell
readings \(F_{rr}\) and \(F_{rf}\), for example. The handlebar length is given
as \(l\) in Equation \ref{eq:angular-impulse}. We use angular impulse instead
of peak torque to normalize for the duration of the applied perturbation to
capture the total effect of the perturbation pulse.

We record the order number of each perturbation \(j\), i.e. first, second,
third, \ldots, to measure how many perturbations a rider is subjected to before
the current perturbation. At the initiation of each perturbation we log the
instantaneous steer and roll angles, \(\delta_0\) and \(\phi_0\), to
characterize the configuration of the bicycle when perturbed. The gain factor
setting on the balance assist controller indicates if the assistance is off
\(\kappa=0\) or on at two different levels: low \(\kappa=8\) or high
\(\kappa=10\). As mentioned earlier, a recovery from the perturbation is
successful if the person neither places their foot down onto the treadmill
surface nor their wheel of the bicycle exits the width of the treadmill belt.
We record this as a binary variable \(f\) for ``fall''.  All measured variables
are reported in Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall Outcome'' and
    ``Perturbation Order Number'' are recorded per perturbation instance. The
    remaining measures are time varying during each trial.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain Factor & \(\kappa\) & \si{\newton\second\squared} & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} &  DF30 wheel encoder \\
    Fall Outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear &
    \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & inline load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & BN0086 rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

\subsection{Statistics}
%
We test our hypothesis that the balance assist control will reduce the
probability of falling when perturbed externally at the handlebar. We have a
single binary fall outcome variable \(f\) that is dependent on several possible
explanatory independent variables, one of which is the binary balance assist
state (on or off). See Table~\ref{tab:stat-model-variables} for the statistical
model variables.
%
\begin{table}
  \centering
  \caption{Independent and dependent variables used in the statistical
    regression model.}
  \label{tab:stat-model-variables}
  \begin{tabular}{llll}
    \toprule
    Variable & Causality & Units & Description \\
    \midrule
    \(L\)        & independent & \si{\newton\meter\second} & angular impulse of perturbation torque \\
    \(\delta_0\) & independent & \si{\degree} & steer angle at start of perturbation \\
    \(\phi_0\)   & independent & \si{\degree} & roll angle at start of perturbation \\
    \(f\)        & dependent   & boolean & outcome: did not fall 0 or did fall 1 \\
    \(s\)        & independent & boolean & balance assist: off 0 or on 1 \\
    \(j\)        & independent & integer & order number of perturbation \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\section{Results}
%
The coefficient estimates for a single-level logistic regression for the data
from the 6~\si{\kph}, gain factor \(\kappa=10\) trials are shown in
Table~\ref{tab:freq-coefs-6}. The angular impulse, perturbation order, and
balance assist state are all statistically significant predictors with angular
impulse having a dominant effect. Larger angular impulse increases the
probability to fall and both enduring more perturbations or having the balance
assist on, decrease the probability to fall. The associated multiplicative
change in odds are also shown in Table~\ref{tab:freq-coefs-6} and can be used
to calculate the probability of falling.
%
\begin{table}
  \centering
  \caption{Logistic regression intercept \(\beta\) and coefficient estimates
    \(\alpha_k\) at 6~\si{\kph} and gain factor \(\kappa=10\) along with the
    standard error SE, p-value \(p\), multiplicative change in odds
    \(e^{\beta}\) or \(e^{\alpha_k}\), and the 5\% confidence interval bounds.}
  \label{tab:freq-coefs-6}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & \(\beta\) / \(\alpha_k\) & SE & \(p\) & \(e^{\beta}\) / \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
    \midrule
    Intercept \(\beta\)                           & -0.29 & 0.17 & 0.09 & 0.75 & 0.53 & 1.05 \\
    Angular impulse \(L\)                         & 1.69 & 0.27 & 0.00\(^*\) & 5.40 & 3.18 & 9.16 \\
    Perturbation order \(j\)                      & -0.77 & 0.22 & 0.00\(^*\) & 0.46 & 0.30 & 0.72 \\
    Balance assist state \(s\)                    & -0.64 & 0.27 & 0.02\(^*\) & 0.53 & 0.31 & 0.89 \\
    Roll angle \(\phi_0\)                         & -0.25 & 0.21 & 0.24 & 0.78 & 0.51 & 1.18 \\
    Steer angle \(\delta_0\)                      & -0.14 & 0.21 & 0.51 & 0.87 & 0.58 & 1.32 \\
    Balance assist state $\times$ roll angle         & 0.52 & 0.34 & 0.12 & 1.68 & 0.86 & 3.29 \\
    Balance assist state $\times$ steer angle        & -0.41 & 0.34 & 0.22 & 0.66 & 0.34 & 1.28 \\
    Balance assist state $\times$ angular impulse    & 0.41 & 0.41 & 0.32 & 1.51 & 0.67 & 3.38 \\
    Balance assist state $\times$ perturbation order & -0.53 & 0.34 & 0.12 & 0.59 & 0.30 & 1.15 \\
    \bottomrule
  \end{tabular}
\end{table}
